Efficiency

### 2018-02-04 DraftExec.tex
with the battery discharge qualification

It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.  In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller". Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.  Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.

@@

In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.

The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.  Not all decelerating ducts have stators, but the impact of the duct on the stream will remain. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.  In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.  However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.  In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.  No definitive transition point seems to be accepted as having great significance.  Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The difference between a pumpjet and a propeller]
In it simplest form, a pumpjet can be thought of as a special type of ducted propeller, or propeller contained inside a shroud.  The impact of the shroud in a pumpjet has the special effect of \textit{decelerating} the water at the point that it reaches the propeller.  This has the dual effect of raising the pressure of the water at the point that the rotor has to do its work on it, which is how pumpjets are able to reduce or eliminate cavitation.
\\
\\
This type of duct shape contrasts with the system most commonly associated with ducted propellers on tug-boats and trawlers, known as accelerating ducts or 'Kort Nozzles'.  This type of duct performs essentially the opposite function to the decelerating duct, contracting in advance of the propeller and acclerating the flow at the stage that it meets propeller.  This effect accentuates the normal contraction of the water-column that occurs in open propellers, and can improve the efficiency of small propellers attempting to exert a large force at low speeds (such as tugs and trawlers) where an optimally sized propeller for such a role would be impractically large for the vessel.  In this circumstance the suction force on the inside of the duct helps generate a positive thrust that is greater than the negative drag on the exterior of the duct.
\end{tcolorbox}

@@

The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.

@@

At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role. The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.  These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.  Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.

\begin{figure}[h]
\captionsetup{width=\textwidth}
\includegraphics[width=\textwidth]{BMTShape.png}
\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface. \parencite{giles2010}}
\label{fig:BMTShape.png}
\end{figure}

@@

The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.  The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.  In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.  This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power. The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form. If the power could be reduced by 5\% (e.g. by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.  As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.

The paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.  At 5kt, the propeller was more than twice as efficient as the waterjet.  This study has been used to directly inform the 'central' efficiency curves assessed in this analysis.

@@

There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs}.  They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed. But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Why pumjet efficiency declines at low speed]
Pumpjets have lower efficiency at lower speeds because the shroud forces the stream of water to flow in a very different pattern to what it would naturally.  By expanding and decelerating the flow at the point where it meets the rotor (where it would contract and accelerate without the duct) the duct forces some of the flows kinetic energy to be converted back to pressure.
\\
\\
This has a terrible impact on efficiency when the incoming water doesn't have very much kinetic energy, and not a great deal more needs to be added to it, as happens when a vessel is moving slowly.  The effect of slowing down water that's already going slowly means that some of the water stops, and in fact turns around and creates a vortex, known as 'flow separation'.  The effective size of the water-colum being accelerated reduces and efficiency declines, as more of the system's energy is absorbed in spinning the water around in circles, rather than accelerating it backwards.
\end{tcolorbox}

@@

Derivations of the 'ideal' efficiencies are frequently made for propellers and jets using basic conservation laws (momentum theory) which consider simple acceleration of some column of water as actuated by a disc representing the impeller or propeller.  A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}. These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the pottential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.  As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.

Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and pumpjets. Whilst I won't repeat the full expansion here, a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended. Since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.   Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.

This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:

@@

The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.  Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.  This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.  This of course isn't practical to build, and also doesn't hold when taken to the etremes the effects of the drag on the propeller blades, or duct, are taken into account.  Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this 'ideal' consideration.

@@

The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".

@@

For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.  This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.  This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.  Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.  This still provides a hard limit on potential improvements that can be attained for a pumpjet that has non-zero duct loss.

@@

\begin{figure}[h] \captionsetup{width=0.6\textwidth}
\includegraphics[width=0.8\textwidth, center]{EfficiencyBulten.png}
\caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).  The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct. Credit: \cite{bulten2006numerical}}
\label{fig:EfficiencyBulten.png}
\end{figure}

@@

\begin{figure}[h]
\captionsetup{width=0.5\textwidth}
\includegraphics[width=0.6\textwidth, center]{PumpUp.png}
\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers. Credit: \cite{stanford2017}}
\label{fig:PumpUp.png}
\end{figure}

@@

\subsection{An assessment of scope for improving the efficiency of a pumpjet at low speeds}
Before continuing to assess the impact of the likely efficiency curves of pumpjets relative to propellers, it is worthwhile considering what the known literature about the theory of pumps, jets, and turbomachinery might tell us about the potential for a radical advance which make might substantially improve the shape of the curve at very low speeds.  Whilst the field of propulsor design is no doubt complex, and I will not attempt to be exhaustive in detailing all of the potential design considerations which might make an improvement.  However, the pursuit of reasonably comprehensive confidence is aided by the the scale of the gap which probably needs to be closed to make pumpjets comparably efficient to propellers at the relevant speeds.  Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.  In other words, this gap will not be closed or noticeably reduced by 'tuning'.  Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems (\cite{bruce1974}, \cite{furuya1988}, \cite{mccormick1963design}, \cite{henderson1964}).  As such, I will only consider those parameters which could plausibly be altered quite substantially, where the consideration of larger changes make the likely limits or trade-offs quite clear.

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Can we just 'tune' the pumpjet to be more efficient at low speeds?]
A effects different effect and factors working in different directions make the process of re-designing a pumpjet quite complex, and bode poorly for a high-efficiency, low-speed pumpjet.  Generally increasing the mass-flow, by increasing duct size, might be assumed to be a first step. However this increases the size of the shroud, and efficiencies can decline further, since the duct inevitably produces some negative thrust.  Moderating the 'deceleration' by making the shape change more gradual could reduce some flow separation, but make the duct much longer, and increase overall friction drag from the surfaces.  Easy wins are not obvious.
\\
\\
Furthermore, with peak efficiencies for pumpjets generally currently well above 25kt, only quite radical design changes could be expected to have the sort of effect that might make a big enough difference (good efficiency below 10kt), which aren't really within the scope of 'tuning'.  The radical redesigns which might look capable of having a significant impact start to look very much like evolving right back to a propeller, which defeats the acoustic purposes of having a pumpjet in the first place.
\end{tcolorbox}

@@

As noted in \cite{henderson1964}, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.  The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.  In this respect, there is "no free lunch" to be had in expanding the size of the intake.  As Henderson et al. conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that "if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}

@@

\begin{figure}[h] \captionsetup{width=0.8\textwidth}
\includegraphics[width=\textwidth]{AcceleratingIdeal.png}
\caption{Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies}
\label{fig:AcceleratingIdeal.png}
\end{figure}

@@

Variations from this central line have been made with consideration for some plausible trade-offs errors, which are intended to be indicative of magnitude possible unknowns in system design and their likely impact.  In the case of the 'high' pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.  The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet. It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.  All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries have been selected by Norway in a selection process which happend almost concurrently with Australia's \parencite{ohff201702}, and would probably have been on offer to Australia if they had been sought. Precise energy density of these batteries is not publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.

@@

>>>>

It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.  In \cite[17]{carlton2007} pump jet is described as an ``interesting development of the classical ducted propeller". Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct ``This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.  Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.

@@

In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that ``the stream of flow through the pumpjet is made to depart from the ``natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.

The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.  Not all decelerating ducts have stators, but the impact of the duct on the stream will remain. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.  In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.  However, there is not to my knowledge any single defining physical characteristic which the transition to the name `impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.  In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.  No definitive transition point seems to be accepted as having great significance.  Most commonly literature will refer to a `rotor', a more general term which could encompass both propellers and impellers.

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The difference between a pumpjet and a propeller]
In it simplest form, a pumpjet can be thought of as a special type of ducted propeller, or propeller contained inside a shroud.  The impact of the shroud in a pumpjet has the special effect of \textit{decelerating} the water at the point that it reaches the propeller.  This has the dual effect of raising the pressure of the water at the point that the rotor has to do its work on it, which is how pumpjets are able to reduce or eliminate cavitation.
\\
\\
This type of duct shape contrasts with the system most commonly associated with ducted propellers on tug-boats and trawlers, known as accelerating ducts or `Kort Nozzles'.  This type of duct performs essentially the opposite function to the decelerating duct, contracting in advance of the propeller and acclerating the flow at the stage that it meets propeller.  This effect accentuates the normal contraction of the water-column that occurs in open propellers, and can improve the efficiency of small propellers attempting to exert a large force at low speeds (such as tugs and trawlers) where an optimally sized propeller for such a role would be impractically large for the vessel.  In this circumstance the suction force on the inside of the duct helps generate a positive thrust that is greater than the negative drag on the exterior of the duct.
\end{tcolorbox}

@@

The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as `submerged waterjets' \parencite{buckingham2008submarine}.

@@

At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for `fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role. The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.  These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.  Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.

\begin{figure}[h]
\captionsetup{width=\textwidth}
\includegraphics[width=\textwidth]{BMTShape.png}
\caption{The `fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface. \parencite{giles2010}}
\label{fig:BMTShape.png}
\end{figure}

@@

The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.  The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.  In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.  This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power. The paper states: ``This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form. If the power could be reduced by 5\% (e.g. by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.  As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.

The paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.  At 5kt, the propeller was more than twice as efficient as the waterjet.  This study has been used to directly inform the `central' efficiency curves assessed in this analysis.

@@

There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs}.  They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed. But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an `ideal' actuator disc start to break down.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Why pumjet efficiency declines at low speed]
Pumpjets have lower efficiency at lower speeds because the shroud forces the stream of water to flow in a very different pattern to what it would naturally.  By expanding and decelerating the flow at the point where it meets the rotor (where it would contract and accelerate without the duct) the duct forces some of the flows kinetic energy to be converted back to pressure.
\\
\\
This has a terrible impact on efficiency when the incoming water doesn't have very much kinetic energy, and not a great deal more needs to be added to it, as happens when a vessel is moving slowly.  The effect of slowing down water that's already going slowly means that some of the water stops, and in fact turns around and creates a vortex, known as `flow separation'.  The effective size of the water-colum being accelerated reduces and efficiency declines, as more of the system's energy is absorbed in spinning the water around in circles, rather than accelerating it backwards.
\end{tcolorbox}

@@

Derivations of the `ideal' efficiencies are frequently made for propellers and jets using basic conservation laws (momentum theory) which consider simple acceleration of some column of water as actuated by a disc representing the impeller or propeller.  A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}. These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the pottential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.  As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.

Such methods are used in \cite[8-14]{wislicenus1973} to construct an `ideal' jet efficiency that applies to waterjets and pumpjets. Whilst I won't repeat the full expansion here, a fundamentally important early conclusion of the derivation is the distinction between `useful work' and the total energy expended. Since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.   Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.

This is expressed in the equation which represents `ideal jet efficiency', which is directly equivalent to `ideal propeller efficiency' using momentum theory of propeller action:

@@

The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.  Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.  This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.  This of course isn't practical to build, and also doesn't hold when taken to the etremes the effects of the drag on the propeller blades, or duct, are taken into account.  Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this `ideal' consideration.

@@

The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: ``Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".

@@

For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.  This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.  This is quite a different constraint to other ``practical" considerations as some have suggested, such as weight, or space constraints.  Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.  This still provides a hard limit on potential improvements that can be attained for a pumpjet that has non-zero duct loss.

@@

\begin{figure}[h] \captionsetup{width=0.6\textwidth}
\includegraphics[width=0.8\textwidth, center]{EfficiencyBulten.png}
\caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).  The impact of increasing duct losses can be clearly seen, as opposed to the `ideal' propeller without a duct. Credit: \cite{bulten2006numerical}}
\label{fig:EfficiencyBulten.png}
\end{figure}

@@

\begin{figure}[h]
\captionsetup{width=0.5\textwidth}
\includegraphics[width=0.6\textwidth, center]{PumpUp.png}
\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of `ideal' pumpjets and propellers. Credit: \cite{stanford2017}}
\label{fig:PumpUp.png}
\end{figure}

@@

\subsection{An assessment of scope for improving the efficiency of a pumpjet at low speeds}
Before continuing to assess the impact of the likely efficiency curves of pumpjets relative to propellers, it is worthwhile considering what the known literature about the theory of pumps, jets, and turbomachinery might tell us about the potential for a radical advance which make might substantially improve the shape of the curve at very low speeds.  Whilst the field of propulsor design is no doubt complex, and I will not attempt to be exhaustive in detailing all of the potential design considerations which might make an improvement.  However, the pursuit of reasonably comprehensive confidence is aided by the the scale of the gap which probably needs to be closed to make pumpjets comparably efficient to propellers at the relevant speeds.  Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.  In other words, this gap will not be closed or noticeably reduced by `tuning'.  Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems (\cite{bruce1974}, \cite{furuya1988}, \cite{mccormick1963design}, \cite{henderson1964}).  As such, I will only consider those parameters which could plausibly be altered quite substantially, where the consideration of larger changes make the likely limits or trade-offs quite clear.

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Can we just `tune' the pumpjet to be more efficient at low speeds?]
Several different factors working in different directions make the process of re-designing a pumpjet quite complex, and bode poorly for a high-efficiency, low-speed redesign.  Generally increasing the mass-flow, by increasing duct size, might be assumed to be a first step. However this increases the size of the shroud, and efficiencies can decline further, since the duct inevitably produces some negative thrust.  Moderating the `deceleration' by making the shape change more gradual could reduce some flow separation, but make the duct much longer, and increase overall friction drag from the surfaces.  Easy wins are not obvious.
\\
\\
Furthermore, with peak efficiencies for pumpjets generally currently well above 25kt, only quite radical design changes could be expected to have the sort of effect that might make a big enough difference (good efficiency below 10kt), which aren't really within the scope of `tuning'.  The radical redesigns which might look capable of having a significant impact start to look very much like evolving right back to a propeller, which defeats the acoustic purposes of having a pumpjet in the first place.
\end{tcolorbox}

@@

As noted in \cite{henderson1964}, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.  The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.  In this respect, there is ``no free lunch" to be had in expanding the size of the intake.  As Henderson et al. conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that ``if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}

@@

\begin{figure}[h] \captionsetup{width=0.8\textwidth}
\includegraphics[width=\textwidth]{AcceleratingIdeal.png}
\caption{Calculations of `ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in `ideal' efficiencies}
\label{fig:AcceleratingIdeal.png}
\end{figure}

@@

Variations from this central line have been made with consideration for some plausible trade-offs errors, which are intended to be indicative of magnitude possible unknowns in system design and their likely impact.  In the case of the `high' pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.  The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet. It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.  All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries have been selected by Norway in a selection process which happend almost concurrently with Australia's \parencite{ohff201702}, and would probably have been on offer to Australia if they had been sought. Precise energy density of these batteries is not readily publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.

@@

Furthermore, batteries of most types tend to have higher levels of internal resistance at higher discharge rates, resulting in energy being lost to heat, which reduces their effective capacity.  As such, there are pracitcal limits to how quickly a batterie's energy can be withdrawn, and some decline in effective capacity as the discharge rate increases.  In the case of lead-acid batteries, the very high rate of discharge associated with the top-speeds of a submarine very substantially reduces their effective capacity.  In the modelling for this paper, no specific consideration for the effects of internal resistance is given, and constant capacity is assumed for all speeds.  Given the that the most substantial results of this paper are at relatively low speeds, perhaps half maxium, and consequently an eigth the maxium discharge rate from the batteries, this will have a negligible effect on the most significant results.  However, the modeled endurances and ranges at top speed may be substantially overstated in practice for a submarine for lead-acid batteries.
